---
title: "Seurat object analysis"
output: html_notebook
---


In this notebook the analysis on the Seurat object of the samples is performed.  
The pipeline creates Seurat objects of the samples.  
These samples, and the needed r packages, are loaded.

```{r}
#loading and installing packages
library(SingleR)
library(Seurat)
library(dplyr)
library(ggsci)
library(DoubletFinder)
library(SeuratData)
library(patchwork)
library(ggplot2)
library(celldex)

#setting the path to the right dir
setwd("~/seurat_analysis")
#loading object files

GEM_sample1_cellranger <- readRDS("GEM_sample1_cellranger_seurat.rds")
GEM_sample2_cellranger <- readRDS("GEM_sample2_cellranger_seurat.rds")
GEM_sample3_cellranger <- readRDS("GEM_sample3_cellranger_seurat.rds")
GEM_sample4_cellranger <- readRDS("GEM_sample4_cellranger_seurat.rds")

```

# Doublet detection
Detection of scRNA works with droplets. There can be a chance that multiple cells are in 1 droplet. This means that some droplets can give information of multiple cells. This is called a doublet detection and these needs to be filtered out before performing the following steps. Samples have to be preprocessed (normalize the data, finding the variable genes, scale the data, run PCA and UMAP) before a doublet detection can be performed.

```{r}
#preprocessing of sample 1
GEM_sample1_filtered = FindVariableFeatures(GEM_sample1_cellranger, verbose = F)
GEM_sample1_filtered = PercentageFeatureSet(GEM_sample1_filtered, "^MT-", col.name = "percent_mito")
GEM_sample1_filtered = ScaleData(GEM_sample1_filtered, vars.to.regress = c("nFeature_RNA", "percent_mito"), 
    verbose = F)
GEM_sample1_filtered = RunPCA(GEM_sample1_filtered, verbose = F, npcs = 20)
GEM_sample1_filtered = RunUMAP(GEM_sample1_filtered, dims = 1:10, verbose = F)
GEM_sample1_filtered <-  NormalizeData(GEM_sample1_filtered)
```

To run the doubletFinder_V3 function, paramSweep_V3 needs to be run. This calculates the best pK value to use in the doubletFinder_V3 function. To calculate nExp the doublet rate estimation is needed. This can be determined using the doublet rate estimation table. In my samples 5000 cells were load so I use 0.023 to calculate nExp.

```{r}

# use paramSweep to define best parameters (pK value) for doublet finder. make barplot to better visualize the mean-variance-normalized bimodality coefficient (BCmvn) per pK value.
sweep.res <- paramSweep_v3(GEM_sample1_filtered) 
sweep.stats <- summarizeSweep(sweep.res, GT = FALSE) 
bcmvn <- find.pK(sweep.stats) 
barplot(bcmvn$BCmetric, names.arg = bcmvn$pK, las=2)
 
#calculate expected number of doublets. doublet rate estimation table is used to calculate nEXP value. 
nExp <- round(ncol(GEM_sample1_filtered) * 0.023) 
GEM_sample1_filtered <- doubletFinder_v3(GEM_sample1_filtered, pN = 0.25, pK = 0.24, nExp = nExp, PCs = 1:10)
```


DoubletFinder is also performed on the other samples.

```{r}

#sample2
GEM_sample2_filtered = FindVariableFeatures(GEM_sample2_cellranger, verbose = F)
GEM_sample2_filtered = PercentageFeatureSet(GEM_sample2_filtered, "^MT-", col.name = "percent_mito")
GEM_sample2_filtered = ScaleData(GEM_sample2_filtered, vars.to.regress = c("nFeature_RNA", "percent_mito"), 
    verbose = F)
GEM_sample2_filtered = RunPCA(GEM_sample2_filtered, verbose = F, npcs = 20)
GEM_sample2_filtered = RunUMAP(GEM_sample2_filtered, dims = 1:10, verbose = F)
GEM_sample2_filtered =  NormalizeData(GEM_sample2_filtered)

sweep.res2 <- paramSweep_v3(GEM_sample2_filtered) 
sweep.stats2 <- summarizeSweep(sweep.res2, GT = FALSE) 
bcmvn2 <- find.pK(sweep.stats2) 
barplot(bcmvn2$BCmetric, names.arg = bcmvn2$pK, las=2)

nExp2 <- round(ncol(GEM_sample2_filtered) * 0.023)  
GEM_sample2_filtered <- doubletFinder_v3(GEM_sample2_filtered, pN = 0.25, pK = 0.08, nExp = nExp2, PCs = 1:10)

#sample3
GEM_sample3_filtered = FindVariableFeatures(GEM_sample3_cellranger, verbose = F)
GEM_sample3_filtered = PercentageFeatureSet(GEM_sample3_filtered, "^MT-", col.name = "percent_mito")
GEM_sample3_filtered = ScaleData(GEM_sample3_filtered, vars.to.regress = c("nFeature_RNA", "percent_mito"), 
    verbose = F)
GEM_sample3_filtered = RunPCA(GEM_sample3_filtered, verbose = F, npcs =20)
GEM_sample3_filtered = RunUMAP(GEM_sample3_filtered, dims = 1:10, verbose = F)
GEM_sample3_filtered =  NormalizeData(GEM_sample3_filtered)

sweep.res3 <- paramSweep_v3(GEM_sample3_filtered) 
sweep.stats3 <- summarizeSweep(sweep.res3, GT = FALSE) 
bcmvn3 <- find.pK(sweep.stats3) 
barplot(bcmvn3$BCmetric, names.arg = bcmvn3$pK, las=2)

nExp3 <- round(ncol(GEM_sample3_filtered) * 0.023)  
GEM_sample3_filtered <- doubletFinder_v3(GEM_sample3_filtered, pN = 0.25, pK = 0.005, nExp = nExp3, PCs = 1:10)

#sample4
GEM_sample4_filtered = FindVariableFeatures(GEM_sample4_cellranger, verbose = F)
GEM_sample4_filtered = PercentageFeatureSet(GEM_sample4_filtered, "^MT-", col.name = "percent_mito")
GEM_sample4_filtered = ScaleData(GEM_sample4_filtered, vars.to.regress = c("nFeature_RNA", "percent_mito"), 
    verbose = F)
GEM_sample4_filtered = RunPCA(GEM_sample4_filtered, verbose = F, npcs =20)
GEM_sample4_filtered = RunUMAP(GEM_sample4_filtered, dims = 1:10, verbose = F)
GEM_sample4_filtered =  NormalizeData(GEM_sample4_filtered)

sweep.res4 <- paramSweep_v3(GEM_sample4_filtered) 
sweep.stats4 <- summarizeSweep(sweep.res4, GT = FALSE) 
bcmvn4 <- find.pK(sweep.stats4) 
barplot(bcmvn4$BCmetric, names.arg = bcmvn4$pK, las=2)

nExp4 <- round(ncol(GEM_sample4_filtered) * 0.023)
GEM_sample4_filtered <- doubletFinder_v3(GEM_sample4_filtered, pN = 0.25, pK = 0.3, nExp = nExp2, PCs = 1:10)


### removal of the bcmvn objects
rm(bcmvn, bcmvn2, bcmvn3, bcmvn4)

```

Remove the doublets from the metadata.

```{r}
#use subset to remove all the doublets from the metadata
#sample1
GEM_sample1_clean <- subset(GEM_sample1_filtered, cells = rownames (GEM_sample1_filtered@meta.data) [which(GEM_sample1_filtered@meta.data$DF.classification == "Singlet")])
#sample2
GEM_sample2_clean <- subset(GEM_sample2_filtered, cells = rownames (GEM_sample2_filtered@meta.data)[which(GEM_sample2_filtered@meta.data$DF.classification == "Singlet")])
#sample3
GEM_sample3_clean <- subset(GEM_sample3_filtered, cells = rownames (GEM_sample3_filtered@meta.data)[which(GEM_sample3_filtered@meta.data$DF.classification == "Singlet")])
#sample4
GEM_sample4_clean <- subset(GEM_sample4_filtered, cells = rownames (GEM_sample4_filtered@meta.data)[which(GEM_sample4_filtered@meta.data$DF.classification == "Singlet")])

#remove redundant objects
rm (sweep.res, sweep.res2, sweep.res3, sweep.res4, sweep.stats, sweep.stats2, sweep.stats3, sweep.stats4)
gc()
```


Merge all the cellranger seurat objects into 1 object.

```{r}
# merging all the seurat objects into 1 object
alldata <- merge(GEM_sample1_clean, c(GEM_sample2_clean,GEM_sample3_clean, GEM_sample4_clean))

#removing the sample objects that are not needed anymore
rm ( GEM_sample1_cellranger, GEM_sample1_count_kallisto, GEM_sample1_genecount_kallisto, GEM_sample2_cellranger, GEM_sample2_count_kallisto, GEM_sample2_genecount_kallisto, GEM_sample3_cellranger, GEM_sample3_count_kallisto, GEM_sample3_genecount_kallisto, GEM_sample4_cellranger, GEM_sample4_count_kallisto, GEM_sample4_genecount_kallisto)

rm (GEM_sample1_filtered, GEM_sample2_filtered, GEM_sample3_filtered, GEM_sample4_filtered, GEM_sample1_clean, GEM_sample2_clean, GEM_sample3_clean, GEM_sample4_clean,)

#running garbage collect to free up memory
gc()

```


# Calculating percentage of features

Calculating the percentage mitochondrial and ribosomal genes per cell and adding it to the metadata. This is done because low-quality or dying cells have high mitochondrial contamination. The percentage of hemoglobin and platelets is calculated to see if there is blood contamination.
These can be filtered out. 
The PercentageFeatureSet() function is used to calculate the percentages.

```{r}
alldata <- PercentageFeatureSet(alldata, "^MT-", col.name = "percent_mito")
alldata <- PercentageFeatureSet(alldata, "^RP[SL]", col.name = "percent_ribo")
alldata <- PercentageFeatureSet(alldata, "^HB[^(P)]", col.name = "percent_hb")
alldata <- PercentageFeatureSet(alldata, "PECAM1|PF4", col.name = "percent_plat")
```

# Making violin plot
A violin plot and a scatterplot is created to visualize the data.
```{r}
#naming the features/ naming the graphs
feats <- c("nFeature_RNA", "nCount_RNA", "percent_mito", "percent_ribo","percent_hb")

#violinplot of alldata
VlnPlot(alldata, group.by = "orig.ident", features = feats, pt.size = 0.1, ncol = 3) + 
    NoLegend()

#scatterplot of alldata. plot 1 shows relation of nCount_RNA with nfeature_RNA
plot1 <-FeatureScatter(alldata, "nCount_RNA", "nFeature_RNA", group.by = "orig.ident", pt.size = 0.5)+ NoLegend()

#plot 2 shows relation of nCount_RNA with percent_mito
plot2 <- FeatureScatter(alldata, "nCount_RNA", "percent_mito", group.by = "orig.ident", pt.size = 0.5)+ NoLegend()

#plot 3 shows relation of nFeature_RNA and percent_mito
plot3 <- FeatureScatter(alldata, "nFeature_RNA", "percent_mito", group.by = "orig.ident", pt.size = 0.5) + NoLegend()

#graph of the 3 scatterplots
plot1 + plot2 +plot3

#violinplot of the percent_mito feature
VlnPlot(alldata, group.by = "orig.ident", features = "percent_mito", pt.size = 0.1, ncol = 1) + 
    NoLegend()
```

It seems that roughly 40% of the transcripts are derived from ribosomal genes and roughly 10% from mitochondrial. 


A plot is created to visualize the total count of the top 10 genes per cell.

```{r}
par(mar = c(4, 8, 2, 1))
C <- alldata@assays$RNA@counts
C <- Matrix::t(Matrix::t(C)/Matrix::colSums(C)) * 100
most_expressed <- order(apply(C, 1, median), decreasing = T)[20:1]
boxplot(as.matrix(t(C[most_expressed, ])), cex = 0.1, las = 1, xlab = "% total count per cell", 
    col = (scales::hue_pal())(20)[20:1], horizontal = TRUE) 

```


# Filtering percent_mito and percent_ribo
  
A cuttoff point has to be chosen to filter out low quality cells. This can be done by using a fixed threshold. By assessing the violin plot, we filter out all cells that have more than 10% mitochondrial counts and have more than 5% ribosomal count. 

```{r}
#select cells with these amount of mito and ribo
selected_mito <- WhichCells(alldata, expression = percent_mito < 10)
selected_ribo <- WhichCells(alldata, expression = percent_ribo > 5)

#subset these from the full object
alldata_filtered <- subset(alldata, cells = selected_mito)
alldata_filtered <- subset(alldata_filtered, cells = selected_ribo)


#violinplot of alldata
VlnPlot(alldata_filtered, group.by = "orig.ident", features = feats, pt.size = 0.1, ncol = 3) + 
    NoLegend()

#scatterplot of filtered data that shows relation of percetn_mito and nCount_RNA
FeatureScatter(alldata_filtered, "nCount_RNA", "percent_mito", group.by = "orig.ident", pt.size = 0.5)

```

# Normalize the data
The next step is to normalize the data. This is done with the NormalizeData() function.

```{r}
alldata_filtered <- NormalizeData(alldata_filtered)
```

# Gene variance
Identify and calculate subset of high cell-to-cell variable genes/features. 

```{r}
#find all the variable genes
alldata_filtered <- FindVariableFeatures(alldata_filtered, selection.method = "vst", nfeatures = 2000)

# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(alldata_filtered), 10)

# plot variable features with and without labels
plot1 <- VariableFeaturePlot(alldata_filtered)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot2
```

The top 10 genes that shows the highest variance are labeled on the graph.

# Cell-cycle score

To mitigate the effects of cell cycle heterogeneity, cell cycle phases are scored and filtered out. The scoring is based on the expression of the markers of the G2/M and S phase of the cell. A list of the markers are build in in Seurat (cc.genes$s.genes/g2m.genes). Seurat has a function, CellCycleScoring(), to calculate this score. A violin plot is made based on this score.
The data is scaled after calculating the score. This is done because highly expressed genes shows the highest variance. The data needs to be scaled so that it scales variation with expression level. 

```{r}
#cc.genes$g2m.genes and cc.genes$s.genes are a list of the markers.
alldata_filtered <- CellCycleScoring(object = alldata_filtered, g2m.features = cc.genes$g2m.genes, s.features = cc.genes$s.genes, set.ident = TRUE)

#scale the data
alldata_filtered <- ScaleData(alldata_filtered, features = rownames(alldata_filtered))

#running PCA on the data 
alldata_filtered <- RunPCA(alldata_filtered, features = VariableFeatures(alldata_filtered))

# creating violin plot based on the cell cycle score
VlnPlot(alldata_filtered, features = c("S.Score", "G2M.Score"), group.by = "orig.ident", 
    ncol = 2, pt.size = 0.1)

# Dim plot of the PCA colored by cell cycle phase
DimPlot(alldata_filtered, reduction = "pca", group.by= "Phase", split.by = "Phase")
```

The plot shows that there is no significant difference in expression based on the cell cycle. Regression of the variation is not needed for these samples.

Save the data so if a prblems occurs, you dont need to rerun these steps.

```{r}
#save data
saveRDS(alldata_filtered, "~/alldata_filtered2")

#load data
alldata_filtered <-readRDS("alldata_filtered2")
```


# Linear dimensional reduction

A PCA is performed with the determined variable features as input. To visualize it a heatmap is created.

```{r}
#running PCA on data
alldata_filtered <- RunPCA(alldata_filtered, features = VariableFeatures(object = alldata_filtered))

# Examine and visualize PCA results a few different ways
print(alldata_filtered[["pca"]], dims = 1:5, nfeatures = 5)

#creating heatmap
DimHeatmap(alldata_filtered, dims = 1:6, cells = 500, balanced = TRUE)

```

With the heatmap you can determine the primary source of heterogeneity in a dataset. The cells and features are ordered according to their PCA score.

# Finding clusters
the clustering is based on the method that embeds the data in a K-nearest neighbor graph. The graph is constructed based on the euclidean distance in PCA space and it refines the edge weights of 2 cells based on their shared overlap in their neighbors. This is done using FindNeighbors() function. then the cells are clustered using FindClusters() function.
The resolution sets the granularity of downstream clusters. resolution between 0.4 and 1.4 is good for this dataset. We will work with resolution 0.8. 

```{r}
#first the neighbors then the clusters are calculated
alldata_filtered <- FindNeighbors(alldata_filtered, dims = 1:10)
 

#choosing resolution 0.8 to form clusters
alldata_filtered <- FindClusters(alldata_filtered, resolution = 0.8)

```

# Non-linear dimensional reduction

2 dimensional reductions, tSNE and UMAP, will be performed on the dataset. This will group similar cells in clusters.

```{r}
# running UMAP on dataset
alldata_filtered<- RunUMAP(alldata_filtered, dims = 1:10)

#creating dimplot
DimPlot(alldata_filtered, reduction = "umap", label = TRUE ,label.size = 6)

#running tSNE on dataset
alldata_filtered <- RunTSNE(alldata_filtered, dims = 1:10)
DimPlot(alldata_filtered, reduction = "tsne")


#save and load object file
saveRDS(alldata_filtered, "~/alldata_filtered_afterUMAP")
alldata_filtered <- readRDS("alldata_filtered_afterUMAP")


```

# Cell type recognition

The cell type is not defined in the object. If we determine the markers per cluster now, it will name each cluster by number. To name each cluster the cell type has to be defined. To define the cell type we use the SingleR() funtion. SingleR() will leverage each cell to a reference to determine its type.

```{r}
#load reference data
hpca.se <- HumanPrimaryCellAtlasData()
hpca.se

#get normalized count from Seurat object
sR_alldata <- GetAssayData(alldata_filtered)

#perfrom SingleR to label markers with cell type
com.res1 <- SingleR(sR_alldata, ref=hpca.se, labels=hpca.se$label.main, assay.type.test=1)
# add the labels from SingleR to the metadata of Seurat object
alldata_filtered[["SingleR.labels"]] <- com.res1$labels

```



# Finding markers

Identifying the markers can help define the clusters. with the FindAllMarkers() funtion all markers will be identified. With the min.pct argument only features, with a minimum of the defined percentage value detected in both clusters, will be accounted for. 

```{r}
#finding all positive markers for each cluster
alldata_filtered2.markers <- FindAllMarkers(alldata_filtered2, only.pos = TRUE, min.pct =  0.25, logfc.threshold = 0.25)

```

A heatmap of the clusters plotted against the markers. 

```{r}
top14 <- alldata_filtered.markers %>%
    group_by(cluster) %>%
    top_n(n = 4, wt = avg_log2FC)
DoHeatmap(alldata_filtered, features = top14$gene) + NoLegend()
```


# Naming the clusters

The markers can be matched to cell types. these then can be projected on the graph to see which cell type each cluster is. The clusters are grouped by the cell type.

```{r}
#create dimplot of the clusters
DimPlot(alldata_filtered, reduction = "umap", group.by = "SingleR.labels", cols= DiscretePalette(n=12, palette = "polychrome"), label = TRUE,   pt.size = 0.5) 

```



